Computing likelihoods under the SMC for a general class of ARGs.

### 2023-11-07 paper.tex
update algorithm description

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts, bbm}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{math,calc,positioning}
\usetikzlibrary{calc,positioning}
\usepackage{subcaption}

@@

More formally, and following the notation of \cite{mcvean_approximating_2005}, at any 
point in time, the process can be described by $L(t)$, the set of labeled lineages 
extant at time $t$ each represented by a union of ordered non-overlapping half open 
intervals detailing the ancestral material 
carried by that lineage $X_i = \{[x_{i0}, y_{i0}), \dotsc, [x_{im}, y_{im})\}$.
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$, each represented by 
single interval spanning the entire genome.
Backwards in time the process evolves by a succession of coalescence and recombination 
events until all segments of ancestral material are each present in only one lineage. 
The waiting time to next event is determined by these two competing processes with exponentially 
distributed waiting times as described below.\\

@@

The process described in the previous paragraph details the transitions of the 
set of labeled lineages $L$ through time. An Ancestral Recombination Graph 
$\mathcal{G} = (N, E)$ is obtained by registering a tuple $(c, p, X_{cp})$ whenever a 
lineage $c$, associated with a disjoint set of genomic intervals $X_c$, is 
replaced by one (coalescence) or two lineage(s) (recombination), $p$ (and $p^{\prime}$).
The last element of the tuple is the extent of the edge. It is itself a  
disjoint set of intervals such that $X_{cp} \cup X_{cp^{\prime}} = X_c$. 
The set of edges $E$, containing all such tuples $(c, p, X_{cp})$, 
and $N$, the set of nodes representing all genomes in the ARG, together
define this data structure.\\

Recording the full process as described above will result in unary nodes.
These are nodes that will have only one child in one or more local trees 
(see fig.\ref{fig:smc-unary} and \citet{Wong-2023}). This is not only true 
for the nodes registered for those lineages hit by a recombination event. 
In case of a coalescence event, a node is registered along the full 
extent of the original lineage $X_c$. 
Coalescence between two lineages without perfectly 
overlapping intervals will therefore result in locally unary nodes.\\

The resulting data structure can be simplified to various degrees \citep{Wong-2023}. 
This equates to removing certain types of unary nodes, 
depending on the degree of information 
we want to retain (or can reconstruct) from the original process. 
Unary nodes therefore are a vital link between 
ARGs and local trees. In a fully simplified ARG, without unary nodes, 
the local trees only store information on local coalescence events.

Here we will compute the likelihood for an ARG
under the SMC for which we have no information on the time to any 
of the recombination events. 
We will assume however, that the identity 
of the lineage involved in each recombination event can be correctly 
inferred by relying on the unary nodes (see section X.X on locally unary nodes).
As highlighted in fig.\ref{fig:smc-unary}, unary nodes can be inferred from the 
available haplotype information. 
Currently, the following methods return this information by default 
as part of their ARG reconstruction algorithm: 
\argweaver \citep{rasmussen_genome_wide_2014} 
\tsinfer \citep{kelleher_inferring_2019}, 
\kwarg \citep{ignatieva_kwarg_2021}.
The presence of these unary nodes enables us to extend likelihood computations 
to the a more general class of ARGs.

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/smc-custom-haplotypes-mini.png}
\caption{SMC with unary nodes: Tree sequence encoding of an ARG (top left) 
generated under the SMC. The haplotypes that make up the lineages coalescing 
at the root of each marginal tree are highlighted at the top. This ancestral 
material is coloured based on the samples that inherit from it.
These reveal the details on the presence of the unary nodes. 
Note how the presence of the unary nodes 
modifies the left-to-right logic of the prune regrafts 
under the SMC (blue dotted arrows).
The unary nodes allow us to keep track of the full extent of all lineages 
through time (under the SMC) and thus infer the lineage along which a 
recombination event happened.}
\label{fig:smc-unary}
\end{figure}

\subsection{Unary nodes} \label{par:unary}

Recording/inferring unary nodes allow us to completely reconstruct 
$X(t)$ through time under the SMC up to the exact 
time of the recombination events. \\

The SMC and unary nodes. \\

inferring recombination = change of parent. \\

@@

\begin{figure}[h!] \label{fig:algo}
\centering
\includegraphics[width=0.9\textwidth]{figures/smc_algo.png}
\caption{Description of algorithm}
\end{figure}

@@

%%%% improve this %%%%%%5
The lineages associated with a particular edge can only coalesce with other lineages 
an edge starting at $x$ can only coalesce 
with a subset of the other edges that make up the local tree at position $x$.
%%%% 
Because local tree changes only occur whenever edges are removed and/or inserted,
we will rely on the ordering of these insertions and removals as implemented in
\texttt{tskit}, to ensure this state can be correctly and efficiently maintained.\\

@@

% improve this section, then transition to limitations of unary nodes
We further rely on the assumption that the (inferrable, see next paragraph) unary nodes 
provide us with the required information on both the extent of all extant lineages through 
time as well as on the time boundaries for any recombination event and the identity of the 
lineage hit by that recombination.

% section on how deviations from the ideal assumptions affect the algorithm
% extent of lineages
% version of the algorithm with polytomies

@@

\bibliographystyle{plainnat}
\bibliography{paper, temp}

>>>>

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{amsmath,amssymb,amsfonts, bbm}%
\usepackage{geometry}%
\usepackage{color}
\usepackage{graphicx}
\usepackage{authblk}
\usepackage{nameref}
\usepackage[right]{lineno}
\usepackage{subcaption}
\usepackage{tikz}

%\usepackage{subcaption}

@@

More formally, and following the notation of \citet{mcvean_approximating_2005}, at any 
point in time, the process can be described by $L(t)$, the set of labeled lineages 
extant at time $t$ each represented by a union of ordered non-overlapping half open 
intervals detailing the ancestral material 
carried by that lineage $X_i = \{[x_{i0}, y_{i0}), \dotsc, [x_{im}, y_{im})\}$.
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$, each represented by 
single interval spanning the entire genome.
Backwards in time the process evolves by a succession of coalescence and recombination 
events until all segments of ancestral material are each present in only one lineage. 
The waiting time to next event is determined by these two competing processes with exponentially 
distributed waiting times as described below.\\

@@

The process described in the previous paragraph details the transitions of the 
set of labeled lineages $L$ through time. An Ancestral Recombination Graph 
$\mathcal{G} = (N, E)$ is obtained by registering a tuple $(c, p, X_{cp})$ whenever a 
lineage $c$, associated with a disjoint set of genomic intervals $X_c$, is 
replaced by one or two lineage(s), $p$ (and $p^{\prime}$), 
in case of a coalescence or recombination respectively.
The last element of the tuple is the extent of the edge. It is itself a  
disjoint set of intervals such that $X_{cp} \cup X_{cp^{\prime}} = X_c$. 
The set of edges $E$, containing all such tuples $(c, p, X_{cp})$, 
and $N$, the set of nodes representing all genomes in the ARG, together
define this data structure.\\

Recording the full process as described above will result in unary nodes.
These are nodes that have only one child in one or more local trees 
(see fig.\ref{fig:smc-unary} and \citet{Wong-2023}). This is not only true 
for the nodes registered for those lineages hit by a recombination event. 
In case of a coalescence event, an edge is registered along the full 
extent of both involved lineages. 
Coalescence between two lineages without perfectly 
overlapping intervals will therefore result in locally unary nodes.\\

These unary nodes ensure the one to one correspondence between an ARG and its 
representation as a sequence of local trees (see fig.\ref{fig:smc-unary}). 
Indeed, the presence of these unary nodes allow us to completely reconstruct 
the state of all extant lineages through time from the local trees.

%under the SMC up to the exact time of the recombination events.

The resulting data structure can be simplified to various degrees \citep{Wong-2023}. 
This equates to removing certain types of unary nodes, 
depending on the degree of information 
we want to retain (or can reconstruct) from the original process. 
In a fully simplified ARG, without unary nodes, 
the local trees only store information on local coalescence events.
Under the SMC, if we only omit recombination nodes, % is there a better word?
we are left with the coalescent nodes, those nodes that are coalescent 
along some fraction of the genome. We can however still identify \emph{all} 
lineages that were hit by a recombination event and constrain the time to 
this event to within a window defined by the age of the node $c$ associated with that 
lineage and the minimum age of the parents of the edges connecting to $c$.

As highlighted in fig.\ref{fig:smc-unary}, (coalescing) unary nodes can be inferred 
from the available haplotype information. Currently, the following methods 
return this information by default as part of their ARG reconstruction algorithm: 
\argweaver \citep{rasmussen_genome-wide_2014} 
\tsinfer \citep{kelleher_inferring_2019}, 
\kwarg \citep{ignatieva_kwarg_2021}.

The presence of these (coalescing) unary nodes enables us to extend likelihood 
computations to a more general class of ARGs.
Here we will compute the likelihood for an ARG
under the SMC for which we have no information on the time to any 
of the recombination events. 
We will assume however, that the identity 
of the lineage involved in each recombination event can be correctly 
inferred by relying on the unary nodes.

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/smc-custom-haplotypes-mini.png}
\caption{ARG (top left) generated under the SMC and its representation as a 
series of local trees. The presence of the unary nodes ensure these two 
representations contain the same information.
The haplotypes that make up the lineages coalescing 
at the root of each marginal tree are highlighted at the top. Ancestral 
material is coloured based on the samples that inherit from it.
Note how the presence of the unary nodes 
modifies the left-to-right logic of single-prune-regrafts induced by 
recombination under the SMC (blue dotted arrows).
Even in the absence of the blue recombination nodes, 
the unary coalescing nodes allow us to reconstruct the state of all lineages 
up to the exact time of the recombination event.
}
\label{fig:smc-unary}
\end{figure}

\subsection{Unary nodes} \label{par:unary}
% Do we need a seperate section on the unary nodes?
% is there more to be said about the SMC inm the presence of unary nodes?
Recording unary nodes has interesting consequences on the left-to-right 
description of the SMC. Looking at a single marginal tree, the presence of 
unary nodes tells us that either \\

@@

\begin{figure}[h!] \label{fig:algo}
\centering
\includegraphics[width=0.8\textwidth]{figures/smc_algo.png}
\caption{Description of algorithm}
\end{figure}

@@

The tree sequence encoding provides us with a natural order on all lineages.  
For each local tree with left coordinate $x$, lineages associated with an edge 
starting at $x$ can only coalesce with the subset of lineages associated with an 
edge that is represented in that same local tree. Because local tree changes 
only occur whenever edges are removed and/or inserted,
we can rely on the ordering of these insertions and removals as implemented in
\texttt{tskit}, to ensure this state can be correctly and efficiently maintained.\\

@@

% improve this section, then transition to limitations of unary nodes
We further rely on the assumption that the unary (coalescent) nodes 
provide us with the required information on both the extent of all extant lineages through 
time as well as on the time boundaries for any recombination event and the identity of the 
lineage hit by that recombination.

% section on how deviations from the ideal assumptions affect the algorithm
However,

% keep track of last parent to identify recombination events
% recombination events beyond root: we keep track of
% parent of prior tree

% extent of lineages

% version of the algorithm with polytomies

@@

\bibliographystyle{plainnat}
\bibliography{paper.bib, temp.bib}
